---
title: "Data Wrangling with dplyr"
---

Business data rarely arrives in the exact shape needed for a decision. Data wrangling is the process of selecting relevant observations, creating useful measures, grouping records, and producing a summary that answers a question.

The dplyr package organizes these tasks into a small set of readable verbs. A pipeline expresses the transformation as a sequence from top to bottom.

## The pipe

The base R pipe, written as |>, passes the result of the expression on its left into the function on its right:

```{r}
#| eval: false
mtcars |>
  head()
```

Many R courses also use the pipe from magrittr, written as %>%. Both styles support the same general idea. The base pipe is part of modern R and avoids requiring a separate package for the pipe itself.

## The core dplyr verbs

### filter(): choose rows

Use filter() to keep observations that satisfy a condition. In a business dataset, this might mean a date range, region, customer segment, or product category.

```{r}
#| eval: false
mtcars |>
  dplyr::filter(cyl >= 6)
```

### select(): choose columns

Use select() to keep only the variables needed for the next step or the final report.

```{r}
#| eval: false
mtcars |>
  dplyr::select(mpg, cyl, hp, wt)
```

### mutate(): create variables

Use mutate() to add or modify columns. Every derived measure should have a clear definition and compatible units.

```{r}
#| eval: false
mtcars |>
  dplyr::mutate(power_to_weight = hp / wt)
```

### summarise(): reduce rows to metrics

Use summarise() to calculate a smaller set of values, often after grouping. Common summary functions include mean(), sum(), median(), min(), max(), and n().

### group_by(): define the comparison

group_by() tells dplyr to calculate summaries separately for each group. A summary without grouping answers a different question from a summary by region, segment, or month.

### arrange(): sort the result

Use arrange() to order rows. Add desc() inside arrange() when the largest values should appear first.

## A complete pipeline with mtcars

The built-in mtcars dataset contains one row per car model. For this teaching example, treat the cars as a product portfolio. We will:

1. keep cars with at least six cylinders;
2. keep the columns needed for the analysis;
3. calculate horsepower per unit of weight;
4. group cars by cylinder count;
5. calculate group-level metrics; and
6. sort the groups by average fuel efficiency.

```{shinylive-r}
#| standalone: true
#| components: [editor, viewer]

library(shiny)
library(dplyr)

car_summary <- mtcars |>
  tibble::as_tibble(rownames = "model") |>
  filter(cyl >= 6) |>
  select(model, mpg, cyl, hp, wt) |>
  mutate(power_to_weight = hp / wt) |>
  group_by(cyl) |>
  summarise(
    cars = n(),
    average_mpg = mean(mpg),
    average_hp = mean(hp),
    average_power_to_weight = mean(power_to_weight),
    .groups = "drop"
  ) |>
  arrange(desc(average_mpg))

ui <- fluidPage(
  h3("Product portfolio summary"),
  p("The table compares car groups with six or more cylinders."),
  tableOutput("summary_table")
)

server <- function(input, output, session) {
  output$summary_table <- renderTable({
    car_summary
  }, digits = 1)
}

shinyApp(ui, server)
```

## Read the pipeline as a business analyst

A pipeline is not automatically correct because it is readable. Ask:

- Does the filter include the population the stakeholder intended?
- Are the selected columns sufficient to audit the calculation?
- Are the units compatible in every mutate() expression?
- Is the grouping level appropriate for the decision?
- Does the summary hide important variation?
- Is the sort order emphasizing the right outcome?

These questions are equally important when the pipeline was written by you or generated by an AI assistant.

