---
title: "R Basics: A Complete Foundation"
---

R is both a programming language and an environment for working with data. A business analyst uses R to represent a question, transform observations, calculate measures, visualize patterns, and create a record of the work.

This chapter builds the vocabulary needed to read and direct R code generated by an AI assistant.

## 1. R as a calculator

R can evaluate arithmetic expressions:

```{r}
#| eval: false
revenue <- 185000
cost <- 112000
profit <- revenue - cost
margin <- profit / revenue
```

The assignment operator <- stores a result under a name. The expression on the right is evaluated first. Use descriptive names that communicate the business meaning of an object.

### Operators

Common operators include:

| Operator | Meaning | Example |
|---|---|---|
| +, -, *, / | Arithmetic | revenue - cost |
| ^ | Exponent | growth_rate ^ 2 |
| == | Equal to | region == "North" |
| != | Not equal to | status != "Cancelled" |
| >, <, >=, <= | Comparison | revenue >= 10000 |
| &, \| | And, or for vectors | active & revenue > 0 |
| ! | Not | !returned |

The single equals sign is used inside function arguments or assignments in other languages, while R commonly uses <- for assignment and == for comparison.

## 2. Core data types

R uses different data types to represent different kinds of information:

| Type | Example | Business use |
|---|---|---|
| Numeric | 185000, 0.12 | Revenue, quantity, margin, rate |
| Character | "North", "Enterprise" | Region, segment, product label |
| Logical | TRUE, FALSE | Flags such as active or approved |
| Integer | 1L, 12L | Counts and positions |
| Date | as.Date("2026-01-15") | Transaction dates and reporting periods |

Inspect an object with class(), typeof(), and str():

```{r}
#| eval: false
amount <- 185000
region <- "North"
approved <- TRUE
report_date <- as.Date("2026-01-15")

class(amount)
typeof(region)
str(list(amount = amount, region = region, approved = approved, report_date = report_date))
```

A type mismatch is a common source of errors. A number stored as character text cannot be reliably added to another number until it is cleaned and converted. A date stored as text may sort alphabetically rather than chronologically.

## 3. Vectors

A vector is a sequence of values. Most columns in a business dataset are vectors, with one value for each observation.

```{r}
#| eval: false
regions <- c("North", "South", "West")
sales <- c(42000, 38500, 51750)
is_target_met <- c(TRUE, FALSE, TRUE)

length(regions)
sales * 1.05
sales > 40000
```

R performs element-by-element operations on vectors. Be careful when vectors have different lengths: R may recycle the shorter vector, which can be useful but can also hide a mistake.

Missing values are represented by NA. Test for them with is.na(); do not compare with == NA.

```{r}
#| eval: false
scores <- c(91, 87, NA, 94)

is.na(scores)
mean(scores, na.rm = TRUE)
```

The business decision about missing data belongs in the analysis plan. Removing missing values can change the population being described.

## 4. Indexing and subsetting

Use square brackets to select positions or conditions:

```{r}
#| eval: false
sales[1]
sales[1:2]
sales[sales > 40000]
regions[sales > 40000]
```

The expression sales > 40000 produces a logical vector. Using that vector inside brackets keeps only the corresponding values. This pattern is the foundation for filtering rows in more advanced tools.

Use the dollar sign to select a named column from a data frame:

```{r}
#| eval: false
regional_sales <- data.frame(
  region = regions,
  sales = sales,
  target_met = is_target_met
)

regional_sales$region
regional_sales[regional_sales$sales > 40000, ]
```

## 5. Data frames and tibbles

A data frame combines equal-length vectors into rows and columns. Each row can represent a transaction, customer, store, or month. Each column represents a variable or measure.

```{r}
#| eval: false
head(regional_sales)
nrow(regional_sales)
ncol(regional_sales)
names(regional_sales)
summary(regional_sales)
```

A tibble is a modern data-frame format used throughout the tidyverse. It prints compactly and makes column behavior easier to inspect. The two structures represent rectangular data; the difference is mostly in printing and modern defaults.

## 6. Functions

A function packages a reusable operation. Functions take arguments and return a result:

```{r}
#| eval: false
mean(sales)
round(mean(sales), digits = 0)
sort(sales, decreasing = TRUE)
```

You can write your own function when a calculation will be repeated:

```{r}
#| eval: false
margin_rate <- function(revenue, cost) {
  (revenue - cost) / revenue
}

margin_rate(185000, 112000)
```

A good business function has clear inputs, a clear output, and a definition that a stakeholder can understand.

## 7. Conditional logic

Use if and else for a decision involving one value. Use ifelse() or dplyr’s if_else() for vectorized decisions.

```{r}
#| eval: false
profit <- 73000

if (profit > 0) {
  "Profitable"
} else {
  "Loss"
}

performance <- ifelse(sales >= 40000, "Above threshold", "Below threshold")
performance
```

Write the threshold and its meaning down. A label such as “high value” is not self-explanatory unless the business rule is documented.

## 8. Repetition and iteration

Many business transformations can be expressed with vectorized operations and dplyr. Base R also provides loops and apply functions for repeated work.

```{r}
#| eval: false
monthly_sales <- c(42000, 38500, 51750)

for (month_sales in monthly_sales) {
  print(month_sales * 1.05)
}
```

Loops are useful, but they are not automatically better. Ask an AI assistant to explain whether a vectorized or grouped solution is clearer before accepting a complex loop.

## 9. Packages, scripts, and projects

R’s core installation is extended by packages. A package may provide functions, datasets, documentation, and dependencies.

- install.packages() downloads a package once;
- library() makes an installed package available in a session;
- package documentation describes intended behavior; and
- a project folder keeps scripts, data, and rendered outputs organized.

A reproducible analysis records the packages and versions it relies on. When an AI-generated script begins with several library() calls, treat them as part of the data contract, not as decoration.

## 10. Practice in a live application

The application below creates objects of several basic types and shows how vectors become a data frame. Run it, then edit the values. Try adding a region, changing a logical flag, or calculating a new margin column.

```{shinylive-r}
#| standalone: true
#| components: [editor, viewer]

library(shiny)

regions <- c("North", "South", "West")
sales <- c(42000, 38500, 51750)
is_target_met <- c(TRUE, FALSE, TRUE)

regional_sales <- data.frame(
  region = regions,
  sales = sales,
  target_met = is_target_met
)

regional_sales$margin <- c(0.31, 0.27, 0.34)

ui <- fluidPage(
  h3("R objects, types, vectors, and data frames"),
  p("The table combines vectors into a small business dataset."),
  h4("Object inspection"),
  verbatimTextOutput("types"),
  h4("Regional sales data"),
  tableOutput("sales_table")
)

server <- function(input, output, session) {
  output$types <- renderPrint({
    list(
      regions_class = class(regions),
      sales_class = class(sales),
      target_flag_class = class(is_target_met),
      table_dimensions = c(nrow(regional_sales), ncol(regional_sales)),
      column_names = names(regional_sales)
    )
  })

  output$sales_table <- renderTable({
    regional_sales
  }, digits = 2)
}

shinyApp(ui, server)
```

## Foundation checklist

Before moving to data wrangling, you should be able to:

- assign and inspect named objects;
- distinguish numeric, character, logical, integer, and date values;
- create and inspect vectors;
- identify and handle missing values;
- subset vectors and data frames;
- read a function call and its arguments;
- explain a conditional rule; and
- identify which packages a script requires.

When an AI assistant proposes code, these foundations let you ask better questions about what the code receives, what it returns, and what the business result means.


