---
title: "Visualization with ggplot2"
---

A good visualization makes a comparison easier to see. It does not replace a definition, a data check, or a well-formed question. Before you choose a chart, identify the variables, the comparison, and the audience.

The ggplot2 package builds plots by combining layers. A typical plot specifies:

- the dataset;
- an aesthetic mapping, such as x, y, color, or fill;
- one or more geometric layers, such as points or bars; and
- labels or a theme that make the result interpretable.

## A scatter plot

A scatter plot is useful when you want to examine the relationship between two numeric variables. In the mtcars data, we can compare vehicle weight with miles per gallon.

```{r}
#| eval: false
library(ggplot2)

ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point() +
  labs(
    x = "Weight, in thousands of pounds",
    y = "Miles per gallon"
  )
```

Each row becomes one point. The mapping in aes() connects columns to visual properties. A trend that appears in a plot is a pattern to investigate, not automatically evidence of causation.

## A bar chart

A bar chart is useful for comparing values across discrete categories. First calculate a summary at the level you want to display; then map the category and measure to the chart.

```{r}
#| eval: false
average_mpg_by_cyl <- aggregate(
  mpg ~ cyl,
  data = mtcars,
  FUN = mean
)

ggplot(average_mpg_by_cyl, aes(x = factor(cyl), y = mpg)) +
  geom_col(fill = "steelblue") +
  labs(
    x = "Number of cylinders",
    y = "Average miles per gallon"
  )
```

A common mistake is to use a bar chart when the x-axis is continuous or to compare totals that have different exposure levels. Always ask whether the denominator and aggregation level make the comparison fair.

## Build both charts interactively

The live application below creates the scatter plot and bar chart. Try changing the x-axis, adding a color mapping, or changing the summary statistic. Notice how a small code change changes the business story the chart can tell.

```{shinylive-r}
#| standalone: true
#| components: [editor, viewer]

library(shiny)
library(ggplot2)

car_data <- mtcars
car_data$model <- rownames(mtcars)
car_data$transmission <- ifelse(car_data$am == 1, "Manual", "Automatic")

average_mpg_by_cyl <- aggregate(
  mpg ~ cyl,
  data = car_data,
  FUN = mean
)

scatter_plot <- ggplot(car_data, aes(x = wt, y = mpg, color = transmission)) +
  geom_point(size = 3) +
  labs(
    title = "Vehicle weight and fuel efficiency",
    x = "Weight, in thousands of pounds",
    y = "Miles per gallon",
    color = "Transmission"
  ) +
  theme_minimal()

bar_plot <- ggplot(average_mpg_by_cyl, aes(x = factor(cyl), y = mpg)) +
  geom_col(fill = "steelblue") +
  labs(
    title = "Average fuel efficiency by cylinder count",
    x = "Cylinders",
    y = "Average miles per gallon"
  ) +
  theme_minimal()

ui <- fluidPage(
  h3("Two views of the mtcars data"),
  plotOutput("scatter"),
  plotOutput("bars")
)

server <- function(input, output, session) {
  output$scatter <- renderPlot(scatter_plot)
  output$bars <- renderPlot(bar_plot)
}

shinyApp(ui, server)
```

## A visualization review checklist

Before sharing a chart, verify:

1. Does the chart match the type of variable and comparison?
2. Are the title, axes, units, and legend understandable without narration?
3. Does the chart use an appropriate denominator?
4. Are unusual values visible and explainable?
5. Could a different scale or aggregation change the conclusion?

An AI tool can generate attractive plotting code quickly. Your job is to check whether the chart represents the data and the decision honestly.

